Thesis Title

### 2023-01-24 chapters/thesis_outline.tex
version 1.0.0
Update Thesis Outline

The work described in the present thesis intends to evaluate the use of bioinformatics methods for the analysis of metagenomic data to allow the rapid identification, virulence analysis and antimicrobial susceptibility prediction of pathogens with clinical relevance. Ultimately, the applicability of metagenomic methods is to be evaluated in a clinical setting as an alternative to current golden standards. Given the dependence of these methodologies on bioinformatics post-processing of the raw data obtained, the major applications and pitfalls of metagenomics are yet to be identified.

The thesis comprises 10 chapters, organised as follows:

@@

@@

\textbf{Chapter 3} describes the application of both second and third-generation sequencing technologies, also known as next-generation and long-read sequencing, to eight tap-water samples collected at the University Medical Center Groningen. Our aim was to evaluate the applicability of shotgun metagenomics, but this time in a surveillance setting. Building on the findings from Chapter 2, a hybrid assembly approach was used to increase resolution power. In this sample a new variant of a colistin resistance (mcr) determinant was detected, named mcr-5.4, and through hybrid assembly leveraging both short and long-read sequences, its context was determined, albeit with questionable success. This chapter is included in the following publication: \textit{G. Fleres, N. Couto, L. Schuele, M. A. Chlebowicz, \underline{C. I. Mendes}, L. W. M. van der Sluis, J. W. A. Rossen, A. W Friedrich, S. García-Cobos, Detection of a novel mcr-5.4 gene variant in hospital tap water by shotgun metagenomic sequencing, Journal of Antimicrobial Chemotherapy, Volume 74, Issue 12, December 2019, Pages 3626–3628.  DOI: \url{https://doi.org/10.1093/jac/dkz363.}}

@@

A key process in metagenomic data analysis is the de novo assembly of raw sequence data since it allows recovering contigs representing the replicons present in the sample, be it genomes, plasmids, or bacteriophages, from a pool of mixed raw reads. \textbf{Chapter 5} employs the same core principles as in Chapter 4, describing a one-stop, user-friendly, containerised, and reproducible workflow, named LMAS, to assess the performance of de novo assembly algorithms for the assembly of second-generation metagenomic sequencing data. The LMAS workflow, which allows users to evaluate performance given a known standard community was implemented in Nextflow, ensuring the transparency and reproducibility of the results obtained. Similarly to Chapter 4, the use of Docker containers provides additional flexibility. The results are presented in an interactive HTML report where global and reference-specific performance metrics can be explored. Currently, 12 de novo assemblers are implemented in LMAS, with the possibility of expansion as novel algorithms are developed.

Despite the advantages of reproducible, containerised workflow, Chapters 4 and 5 still do not guarantee the interoperability of results obtained from various sources. Chapter 5 highlighted the impact that the tool choice can have on downstream results when working with metagenomic data, therefore, and due to the lack of standardisation, it is pivotal that results from various tools can be compared for their applicability in the clinic. With a focus on antimicrobial resistance, \textbf{Chapter 6} presents a standardised output specification for the bioinformatic detection of antimicrobial resistance directly from genomes or metagenomes. This addresses the problem of combining the outputs of disparate antimicrobial resistance gene detection tools into a single unified format, implemented into a python package and command-line utility hAMRonization. As the detection of antimicrobial resistance directly from genomic or metagenomic data has become a standard procedure in public health, with hAMRonization allowing for the comparison of results within bioinformatics workflows, as these tools, although implementing similar principles, differ in supported inputs, search algorithms, parameterisation, and underlying reference databases.

@@

\textbf{Chapter 8} showcases an effort to raise standards on the development and distribution of code for bioinformatic analysis. For this, seven recommendations are presented that help researchers implement software testing in microbial bioinformatics. We propose collaborative software testing as an opportunity to continuously engage software users, developers, and students to unify scientific work across domains. As automated software testing remains underused in scientific software, our set of recommendations not only ensures that appropriate effort can be invested in producing high-quality and robust software but also increases engagement in its sustainability. This chapter is included in the following publication: \textit{B. C. L. van der Putten$^*$, \underline{C. I. Mendes}\footnote[1]{These authors contributed equally to this work.}, B. M. Talbot, J. de Korne-Elenbaas, R. Mamede, P. Vila-Cerqueira, L. P. Coelho, C. A. Gulvik, L. S. Katz, The Asm Ngs Hackathon Participants, Software testing in microbial bioinformatics: a call to action. Microbial Genomics, Volume 8, Issue 3. DOI: \url{https://doi.org/10.1099/mgen.0.000790.}}

@@

\textbf{Chapter 10} Contains the main conclusions driven from this work and also perspectives for future work.

>>>>

The work described in the present thesis intends to evaluate the use of bioinformatics methods for the analysis of metagenomic data to allow the rapid identification, virulence analysis and antimicrobial susceptibility prediction of pathogens with clinical relevance. Ultimately, the applicability of metagenomic methods is to be evaluated in a clinical setting as an alternative to current gold standards. Given the dependence of these methodologies on bioinformatics post-processing of the raw data obtained, the major applications and pitfalls of metagenomics are yet to be identified.

The thesis organisation reflects the six publications included, with each of the results chapters containing mostly the content of the publication indicated at the start of that chapter. Given this option, in addition to an introductory preface giving the context of the work performed, each results chapter has its own introduction, materials and methods, discussion, and conclusion. Overall, the thesis comprises 10 chapters, organised as follows: The first, second to last and last chapters include the general introduction, general discussion and main conclusions of the work developed in this thesis, respectively; while the body of the thesis work, corresponding to chapters 2 to 8, is organised into four parts aggregating chapters under a broader topic, numbered from I to IV.

@@

\textbf{Part I} of the results, named “\textit{Applying metagenomics in the clinical context}”, includes \textbf{Chapters 2, 3 and 4}, where metagenomics is applied to both clinical and environmental samples. In the two first chapters, the performance of metagenomics to identify pathogens and antimicrobial resistance genes was evaluated. Importantly it was recognized that there was a lack of standardisation in the bioinformatic metagenomics analysis, so a fully reproducible workflow was developed for the streamlined analysis of human samples for genomic characterization in the context of Dengue infections, as a model for other applications.

@@

\textbf{Chapter 3} describes the application of both second and third-generation sequencing technologies, also known as next-generation and long-read sequencing, to eight tap-water samples collected at the University Medical Center Groningen. Our aim was to evaluate the applicability of shotgun metagenomics, but this time in a surveillance setting. Building on the findings from Chapter 2, a hybrid assembly approach was used to increase resolution power. In this sample a new variant of a colistin resistance (mcr) determinant was detected, named \textit{mcr-5.4}, and through hybrid assembly leveraging both short and long-read sequences, its context was determined, albeit with questionable success. This chapter is included in the following publication: \textit{G. Fleres, N. Couto, L. Schuele, M. A. Chlebowicz, \underline{C. I. Mendes}, L. W. M. van der Sluis, J. W. A. Rossen, A. W Friedrich, S. García-Cobos, Detection of a novel mcr-5.4 gene variant in hospital tap water by shotgun metagenomic sequencing, Journal of Antimicrobial Chemotherapy, Volume 74, Issue 12, December 2019, Pages 3626–3628.  DOI: \url{https://doi.org/10.1093/jac/dkz363.}}

@@

\textbf{Part II} includes only \textbf{Chapter 5}, and is named “\textit{The impact of de novo assemblers in metagenomics}”. A key process in metagenomic data analysis is the de novo assembly of raw sequence data since it allows recovering contigs representing the replicons present in the sample, be it genomes, plasmids, or bacteriophages, from a pool of mixed raw reads. \textbf{Chapter 5} employs the same core principles as in Chapter 4, describing a one-stop, user-friendly, containerised, and reproducible workflow, named LMAS, to assess the performance of de novo assembly algorithms for the assembly of second-generation metagenomic sequencing data. The LMAS workflow, which allows users to evaluate performance given a known standard community was implemented in Nextflow, ensuring the transparency and reproducibility of the results obtained. Similarly to Chapter 4, the use of Docker containers provides additional flexibility. The results are presented in an interactive HTML report where global and reference-specific performance metrics can be explored. Currently, 12 de novo assemblers are implemented in LMAS, with the possibility of expansion as novel algorithms are developed. This chapter is included in the following publication: \textit{\underline{C. I. Mendes}, P. Vila-Cerqueira, Y. Motro, J. Moran-Gilad, J. A. Carriço, M. Ramirez. LMAS: Evaluating metagenomic short de novo assembly methods through defined communities. GigaScience, Volume 12, 2023, giac122, \url{https://doi.org/10.1093/gigascience/giac122}}.

Not only is the analysis of the data critical, but also the post-processing and its contextualisation with other relevant available metadata, to make the most of genomics in the context of clinical microbiology, infection control and public health. \textbf{Part III}, covering \textbf{Chapters 6 and 7}, is focused on the “\textit{Challenges of data availability in metagenomics and beyond}”. Despite the advantages of reproducible, containerised workflow, Chapters 4 and 5 still do not guarantee the interoperability of results obtained from various sources. Chapter 5 highlighted the impact that the tool choice can have on downstream results when working with metagenomic data, therefore, and due to the lack of standardisation, it is pivotal that results from various tools can be compared for their applicability in the clinic. With a focus on antimicrobial resistance, \textbf{Chapter 6} presents a standardised output specification for the bioinformatic detection of antimicrobial resistance directly from genomes or metagenomes. This addresses the problem of combining the outputs of disparate antimicrobial resistance gene detection tools into a single unified format, implemented into a python package and command-line utility hAMRonization. As the detection of antimicrobial resistance directly from genomic or metagenomic data has become a standard procedure in public health, with hAMRonization allowing for the comparison of results within bioinformatics workflows, as these tools, although implementing similar principles, differ in supported inputs, search algorithms, parameterisation, and underlying reference databases.

@@

\textbf{Part IV} includes only \textbf{Chapter 8}, and focuses on the use of “\textit{Crowdsourcing to improve software robustness in metagenomics and beyond}”. \textbf{Chapter 8} showcases an effort to raise standards on the development and distribution of code for bioinformatic analysis. For this, seven recommendations are presented that help researchers implement software testing in microbial bioinformatics. We propose collaborative software testing as an opportunity to continuously engage software users, developers, and students to unify scientific work across domains. As automated software testing remains underused in scientific software, our set of recommendations not only ensures that appropriate effort can be invested in producing high-quality and robust software but also increases engagement in its sustainability. This chapter is included in the following publication: \textit{B. C. L. van der Putten$^*$, \underline{C. I. Mendes}\footnote[1]{These authors contributed equally to this work.}, B. M. Talbot, J. de Korne-Elenbaas, R. Mamede, P. Vila-Cerqueira, L. P. Coelho, C. A. Gulvik, L. S. Katz, The Asm Ngs Hackathon Participants, Software testing in microbial bioinformatics: a call to action. Microbial Genomics, Volume 8, Issue 3. DOI: \url{https://doi.org/10.1099/mgen.0.000790.}}

@@

\textbf{Chapter 10} contains the main conclusions derived from this work and also perspectives for future work.
